### 2023-02-07 article.tex
Improve

\paragraph{Modern forms of paradata}

A recent paper explores the use of paradata to increase the impact of data,
stating that lack of paradata can be seen as 'a drastic constraint'
in the use of data and offer some suggestions to 
make paradata useful for data re(use):
paradata should be comprehensive, documented in a useful 
way (and shaped by the data) and be friendly to computers \cite{huvila2022improving}.
These suggestions appear to work with code as well.
The lack of computer code is just as much 'a drastic constraint'
in the use of a (computational) scientific experiment.
The reasons why paradata in the form of code matters, 
are the same reaons as why paradata matters,
which are data reuse, research reproducibility, 
contextual understanding of disciplinary differences,
Open Science and insights into 
(scientific) knowledge production \cite{huvila2022improving}.
This chapter will focus mostly on the goals of making science
reproducible and code reusable.

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=\linewidth]{figure_1.png}
  \caption{
    Left: general relation between data, paradata and metadata.
    Middle: the same relations specified for eBird.
    Right: the same relations specified for genetic epidemiology.
  }
  \label{fig:figure_1}
\end{figure}

@@

% \paragraph{Introduction}

@@

% \paragraph{Publication of code}

Going back to the example scientists in the introduction,
where one does and the other does not publish the code,
it is not common to publish the code of an experiment or analysis 
\cite{stodden2011trust,read2015sizing} (with a pleasant exception 
being \cite{conesa2019making}).
For example, in computer graphics, 
a field intimately familiar with computer code,
only 192 of 454 SIGGRAPH papers supply computer code \cite{bonneel2020code}.
Another study analysed the reproducibility of registered reports
in the field of psychology, 
where still only 37 out of 62 studies supplied the code 
to redo the analysis \cite{obels2020analysis}.

% \paragraph{For some fields, the experiment is actually run by code}

@@

% \paragraph{The experiments within genetic epidemology works are done by code}

@@

The code for this experiment is vital paradata.
Like all paradata, its omission would be a constraint in the
use of the conclusion reached by it:
we can use it to assess the trustworthiness of the finding,
for example, by running the code again.
Ideally, the code is comprehensive, comprehensively documented
and be friendly to computers \cite{huvila2022improving}.
Code being friendly to computer seems like an obvious 'yes',
yet this is naive, as commonly it is hard to run code in practice. 
% TODO: Link to correct paragraphs
For code to be called computer-friendly, it must be run on an independent entity, 
as is discussed in the paragraphs about hosting code and/or putting the code in a container.
% TODO: Link to correct paragraphs
Also note that there is a problem due to the inherent sensitivity of the genetic data,
which is discussed in paragraph.
Lastly, note that the researcher used data that was itself generated by code:
complex bioinformatics pipelines are used to obtain reliable DNA sequences 
from the biological sample.

% \paragraph{Science should be reproducible}

A scientific claim must be reproducible before it is accepted
by the scientific community.
Reproducibility (i.e. to reproduce the same results) 
and replicability (i.e. to reconclude a conclusion)
are fundamental characteristics of scientific studies \cite{patil2019visual}.
For computational science, theoretically, it should be relatively easy to 
reproduce an experiment, as all it takes is a computer, electricity,
an optional internet connection, the code and the data.
In practice, however, it appears that 
the academic culture to reproduce results 
has been lost over time in general \cite{peng2011reproducible},
with labs that embrace reproducibility (for example, \cite{barba2016hard})
are the exception.

That same study recommended, as a solution to counter this, 
to make reproducible research a minimal requirement for 
publication \cite{peng2011reproducible}.

% \paragraph{Code should be published}

@@

The code of a computational experiment must be published
for an experiment to be easily reproducible.
It is however (too) common that the original data
and/or code are unavailable \cite{peng2021reproducible}.
One saddening example is described in \cite{haibe2020importance}, where
an algorithm that detects breast cancer from images better than a human expert,
was irreproducible.
In the case of genetic epidemiology, 
the original data usually cannot be published, 
due to the guaranteed privacy of the people having shared their characteristics.
The code of a computational experiment, however, \emph{can} be
distributed without such problems.
In academic journals, however, its is not code that is published, 
but an English description of what it does instead.

% \paragraph{Example of code being the ground truth}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Sensitive data}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\paragraph{Making paradata useful}

A recent paper explores the use of paradata to increase the impact of data,
stating that lack of paradata can be seen as 'a drastic constraint'
in the use of data and offer some suggestions to 
make paradata useful for data re(use):
paradata should be comprehensive, documented in a useful 
way (and shaped by the data) and be friendly to computers \cite{huvila2022improving}.
These suggestions appear to work well for code too
and will be discussed in detail below.

\paragraph{The goal of good paradata}

The reasons why useful paradata matters,
are data reuse, research reproducibility, 
contextual understanding of disciplinary differences,
Open Science and insights into 
(scientific) knowledge production \cite{huvila2022improving}.
This chapter discusses these reasons below.

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=\linewidth]{figure_1.png}
  \caption{
    Left: general relation between data, paradata and metadata.
    Right: the same relations specified for genetic epidemiology.
  }
  \label{fig:figure_1}
\end{figure}

@@

\paragraph{Introduction}

@@

\paragraph{How commonly code is published}\label{para:how-commonly-code-is-published}

Going back to the example scientists in the introduction,
where one does and the other does not publish the code,
it is not common to publish the code of an experiment or analysis 
\cite{stodden2011trust,read2015sizing} (with a pleasant exception 
being \cite{conesa2019making}).
For example, in computer graphics, 
a field intimately familiar with computer code,
only 192 of 454 SIGGRAPH papers supply computer code \cite{bonneel2020code}.
Another study analysed the reproducibility of registered reports
in the field of psychology, 
where still only 37 out of 62 studies supplied the code 
to redo the analysis \cite{obels2020analysis}.
Unpublished code has been the cause of some saddening examples,
such as an algorithm that detects breast cancer from images 
better than a human expert \cite{haibe2020importance}.

\paragraph{For some fields, the experiment is actually run by code}

@@

\paragraph{The experiments within genetic epidemology works are done by code}

@@

\paragraph{Code is vital}

The code for this experiment is vital paradata.
Ideally, the code is useful paradata, 
as discussed in section \label{sec:making-code-useful-paradata}.

\paragraph{The goal of useful paradata is reproducible science}

The reason behind paradata is, among others,
to assure reproducible research \cite{huvila2022improving}.
Reproducibility (i.e. to reproduce the same results) 
and replicability (i.e. to reconclude a conclusion)
are fundamental characteristics of scientific studies \cite{patil2019visual}.
For computational science, theoretically, it should be relatively easy to 
reproduce an experiment, as all it takes is a computer, electricity,
an optional internet connection, the code and the data.
In practice, however, it appears that 
the academic culture to reproduce results 
has been lost over time in general \cite{peng2011reproducible},
with labs that embrace reproducibility (for example, \cite{barba2016hard})
are the exception.
To counter the lack of published code, 
it has been suggested to make the reproduction of research 
a minimal requirement for publication \cite{peng2011reproducible}.

\paragraph{Sensitive data}

A genetic epidemiology works with sensitive data as well:
the genetic sequences of participants are private.
For research to be reproducible one needs both the code and the data
to reproduce the (hopefully) same results.
This problem is discussed in section \ref{sec:sensitive-data}.

\paragraph{Code holds the ground truth}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Sensitive data}\label{sec:sensitive-data}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
